---
title: "Single mouse RNA-seq analysis"
author: "Kristen Wells"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: "tango"
    df_print: "paged"
    code_folding: "hide"
    self_contained: true
params:
  genome: "GRCm38"
  DE_alpha: 0.05
  DE_lfc: 1
  counts: "results/red_green_one_mouse_countTable_cutadapt_trim.txt"
  fastqc_one: "results/fastqc_pre_trim_summary_untrimmed.tsv"
  fastqc_two: "results/fastqc_cutadapt_summary_trimmed.tsv"
  star_stats: "results/star_summary_cutadapt_trim.tsv"
  output_dir: "results/R_analysis"
  sample_info: "files/sample_info.csv"
  sample_column: "group"
  fastq_screen: "results/fastq_screen/fastq_screen_info.csv"
  snakemake: FALSE
  genes: 
    - "Mafa"
    - "Pcsk1"
    - "G6pc2"
    - "Syt5"
    - "Syt10"
    - "Igf1r"
    - "Nr4a1"
    - "Nr4a2"
    - "Nr4a3"
    - "Rasd1"
    - "Prkar1a"
    - "Kcnt1"
    - "Piezo1"
    - "Pyy"
    - "Ptger3"
    - "Hk1"
    - "Sik1"
    - "Nrxn2"
    - "Ins2"
    - "Pdx1"
    - "Gck"
    - "Hk2"
    - "Pkm"
    - "Kcnj11"
    - "Abcc8"
    - "Fos"
    - "Gjd2"
    - "Glp1r"
    - "Gcg"
    - "Sst"
    - "Ldha"
    - "Kdm1a"
    - "Irs2"
---


```{r "Load packages and set options", include = F}
library(knitr)
library(cowplot)
library(tidyverse)
library(Matrix)
library(viridis)
library(RColorBrewer)
library(openxlsx)
library(DESeq2)
library(LaCroixColoR)
library(here)
library(MetBrewer)
library(pheatmap)
library(gprofiler2)

# Set default chunk options
opts_chunk$set(
  message = F, 
  warning = F,
  comment = ""
)

```


```{r "Load Theme"}
source(here("src/scripts/functions.R"))

################
# Theme colors #
################

# Colors for heatmap (from the ArchR package)
blueYellow <- c("#352A86", "#343DAE", "#0262E0", "#1389D2", "#2DB7A3",
	"#A5BE6A", "#F8BA43", "#F6DA23", "#F8FA0D")

# Colors for statistics
fastqc_colors <- c("#e31a1c", "#238443", "#ec7014")

star_colors <- c("#737373", "#8c6bb1", "#ec7014",
                 "#e31a1c", "#238443", "#225ea8")

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r load-data}
run_GO <- TRUE
if(params$snakemake){
 # Read in out dir
  output_directory <- params$output_dir
  
  # Read in count table
  count_table <- read.table(params$counts, sep = "\t",
                            row.names = 1, header = T)
  
  # Read in the sample table
  sample_table <- read.table(params$sample_info, sep = ",",
                             header = T)
  
  # Read in star stats
  star_stats <- read.table(params$star_stats, sep = "\t", row.names = 1,
                           header = T)
  
  # Load in fastqc results
  fastqc_summary_one <- read.table(params$fastqc_one, sep = "\t")
  
  # Load in fastqc results
  fastqc_summary_two <- read.table(params$fastqc_two, sep = "\t")

  fastq_screen_res <- read.csv(params$fastq_screen)

  

} else {
  # Read in out dir
  output_directory <- here(params$output_dir)
  
  # Read in count table
  count_table <- read.table(here(params$counts), sep = "\t",
                            row.names = 1, header = T)
  
  # Read in the sample table
  sample_table <- read.table(here(params$sample_info), sep = ",",
                             header = T)
  
  # Read in star stats
  star_stats <- read.table(here(params$star_stats), sep = "\t", row.names = 1,
                           header = T)
  
  # Load in fastqc results
  fastqc_summary_one <- read.table(here(params$fastqc_one), sep = "\t")
  
  # Load in fastqc results
  fastqc_summary_two <- read.table(here(params$fastqc_two), sep = "\t")

  fastq_screen_res <- read.csv(here(params$fastq_screen))

}
```

## Background
This document goes through the analysis of the  Red/Green experiment with Beta Cells.

```{r, "Load in and set up data"}
# Fix up files

################
# Sample table #
################

rownames(sample_table) <- sample_table$Sample_name
rownames(sample_table) <- make.names(rownames(sample_table))

# Remove any samples that weren't in the count table
sample_table <- sample_table[(rownames(sample_table) %in%
                                 colnames(count_table)),]

# Remove failed sample
# Failed was M 4
sample_table <- sample_table %>%
  dplyr::filter(mouse != "4")

# Ensure the count table and sample table are in the same order
count_table <- count_table[ , rownames(sample_table)]

########
# Star #
########

# Filter for mapping % and total read counts
star_stats <- star_stats[grepl(
  "%|Uniquely mapped reads|mapped to multiple loci",
  rownames(star_stats)), ]

star_stats <- star_stats[!grepl(
  "^Mismatch|chimeric", rownames(star_stats)), ]

# Calculate median size
star_stats_median <- star_stats[grepl(
  "Uniquely mapped reads", rownames(star_stats)), ]

median_size <- median(as.numeric(
  star_stats_median["Uniquely mapped reads number", ])) / 1000000

median_rate <- median(as.numeric(str_remove(
  star_stats_median["Uniquely mapped reads %", ], "%")))

alignment_qual <- get_alignment_qual(median_rate)

# Pull out all sample names
samples <- colnames(star_stats)

# Make a column of metrics
star_stats$metric <- rownames(star_stats)

# Make into long form for ggplot
star_stats_long <- gather(star_stats, key = "Sample", value = "value",
                          all_of(samples))


# Add column for value type
star_stats_long <- star_stats_long %>%
  mutate(
    value = as.numeric(str_remove(value, "%")),
    val_type = ifelse(grepl("%", metric),
                      "pct", "int")
  )

# Pull out read counts
# Calculate median library size
read_counts <- star_stats_long %>%
  filter(grepl("mapped reads number", metric)) %>%
  mutate(value = round(value / 1000000)) %>%
  mutate(value = str_c(value, " million"))

##########
# FastQC #
##########

# Change the colnames
colnames(fastqc_summary_one) <- c("Result", "Test", "Sample")

# Update sample names
fastqc_summary_one$Sample <- sub("_S[0-9]*_", "",
                                 fastqc_summary_one$Sample)
fastqc_summary_one$Sample <- sub("_001.fastq.gz", "",
                                 fastqc_summary_one$Sample)

# Change the colnames
colnames(fastqc_summary_two) <- c("Result", "Test", "Sample")

# Update sample names
fastqc_summary_two$Sample <- sub("_trimmed.fastq.gz", "",
                                 fastqc_summary_two$Sample)
fastqc_summary_two$Sample <- sub("_trimmed.fastq.gz", "",
                                 fastqc_summary_two$Sample)

##############
# Set colors #
##############

# Sample
color_builder <- grDevices::colorRampPalette(colors = RColorBrewer::brewer.pal(n = 9, name = "Set1"))
sample_colors <- color_builder(nrow(sample_table))

names(sample_colors) <- rownames(sample_table)

# Cassatt 1
sex_colors <- met.brewer("Cassatt1",
                               length(unique(sample_table$sex)),
                               type="continuous") %>%
  as.character()

names(sex_colors) <- unique(sample_table$sex)

# Egypt for mutant/control
group_colors <- met.brewer("Egypt", 
                           n = 4,
                           type = "continuous") %>%
  as.character()
group_colors <- group_colors[c(1,3)]
names(group_colors) <- unique(sample_table$group)

# Derain for batch 1
mouse_colors <- met.brewer("Derain",
                              n = length(unique(sample_table$mouse)),
                              type = "continuous")
names(mouse_colors) <- unique(sample_table$mouse)

```

# Quality Control {.tabset}

## FastQC Summary
`FastQC` was used to assess the quality of each fastq file. A summary of the results is shown below. Overall, this is okay. I generally focus most on the per tile sequence quality, the per sequence quality scores, the per base sequence quality, sequence length distribution, and the adapter content. All of your sequencing quality look good. The adapter content fails, but I trimmed the adapters before aligning.

There is also failures in the sequence duplication levels. I will check these out in the fastqc output, but this is pretty common with RNA-seq libraries.
```{r "Create fastqc summary", fig.width = 8.5, fig.height = 5}
fastqc_plot <- ggplot(data = fastqc_summary_one,
                      mapping = aes(x = Sample,
                                    y = Test,
                                    fill = Result)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_manual(values = fastqc_colors) +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 8, color = "black"),
    axis.title   = element_blank(),
    axis.text    = element_text(size = 8, color = "black"),
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

fastqc_plot
```

## FastQC after trimming Summary
Here I've reapeated `Fastqc` after adapter trimming. You can ignore most of it, but I wanted to include this so that you could see that the adapters were successfully removed by my trimming with `cutadapt` (the bottom row).
```{r "Create fastqc summary2", fig.width = 8.5, fig.height = 5}
fastqc_plot <- ggplot(data = fastqc_summary_two,
                      mapping = aes(x = Sample,
                                    y = Test,
                                    fill = Result)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_manual(values = fastqc_colors) +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 8, color = "black"),
    axis.title   = element_blank(),
    axis.text    = element_text(size = 8, color = "black"),
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

fastqc_plot
```

## Alignment Summary
Reads were aligned to the `r params$genome` genome using the `STAR` RNA-seq aligner. A summary of the results is shown below, the library size is displayed on each bar. The median library size was `r median_size` million reads and the median alignment rate was `r median_rate`. Overall, the alignment rate was **`r alignment_qual`**. 

This alignment looks quite good to me. The percent mapping is pretty consistent between samples and is quite high.

```{r "Create alignment summary", fig.width = 8.5, fig.height = 5}

# Plot alignment stats
star_stats_long_pct <- star_stats_long %>%
  filter(val_type == "pct") %>%
  
  # Add median rates to metric label
  mutate(
    metric = str_remove(metric, "% of reads | reads %"),
    metric = str_to_sentence(metric)
  ) %>%
  arrange(value) %>%
  mutate(metric = fct_inorder(metric))
  
# Plot STAR alignment stats
star_plot <- ggplot(data = star_stats_long_pct,
                    mapping = aes(x = Sample,
                                  y = value,
                                  fill = metric)) +
  geom_bar(stat = "identity", color = "white", size = 0.5) +
  
  # Label bars with library size
  geom_text(
    data = read_counts, 
    aes(Sample, 50, label = value),
    show.legend = F,
    inherit.aes = F,
    color = "white", 
    angle = 90
  ) +
  
  scale_fill_manual(
    values = star_colors,
    guide  = guide_legend(reverse = T)
  ) +
  scale_y_continuous(labels = function(x) {str_c(x, "%")}) +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 8, color = "black"),
    axis.title   = element_blank(),
    axis.text    = element_text(size = 8, color = "black"),
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

star_plot

```

## Fastq screen
I also ran the alignment against many different genomes to ensure any unmapped reads could be mapped to any type of contamination. I did this using `fastq_screen`

```{r}

# Make the plotting levels better?
fastq_screen_res$Genome <- factor(fastq_screen_res$Genome,
                                  levels = unique(fastq_screen_res$Genome))

# Pick some colors
colors <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(name = "Set1", n = 9))(length(levels(fastq_screen_res$Genome)))

names(colors) <- levels(fastq_screen_res$Genome)

# Make a barplot
barplot <- ggplot2::ggplot(fastq_screen_res, ggplot2::aes(x = sample,
                                                          y = percent, 
                                                          fill = Genome)) +
	ggplot2::geom_bar(stat = "identity", position = "stack") +
	ggplot2::coord_flip() +
	ggplot2::scale_fill_manual(values = colors)

print(barplot)

```


## PCA
Principal component analysis was performed to assess similarities between samples. Here I've made many plots to look at different technical effects in your data. 

A) PCA colored by sample
B) PCA colored by group
C) PCA colored by sex
D) PCA colored by mouse

```{r "Run DESeq2 and create PCA plot", fig.width = 10, fig.height = 8}
# This makes a DESeq2 object where the count data is our count matrix and the colData is our sample data. I am currently making the design based on "group" but we can change this if necessary
sample_table$Sample_name <- factor(sample_table$Sample_name)
sample_table$genotype <- factor(sample_table$genotype)
sample_table$treatment <- factor(sample_table$treatment)
sample_table$group <- factor(sample_table$group)
sample_table$sex <- factor(sample_table$sex)
sample_table$mouse <- factor(sample_table$mouse)

sample_table <- sample_table[order(sample_table$mouse),]

sample_table$mouse2 <- factor(c(1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 4, 4, 5, 5, 3, 3))
count_table <- count_table[ , rownames(sample_table)]

# Set model matrix
mm <- model.matrix(~sex + ~sex:mouse2 + ~sex:group, sample_table)

# Remove zeros
all.zero <- apply(mm, 2, function(x) all(x==0))
all.zero
idx <- which(all.zero)
mm <- mm[,-idx]
dds <- DESeqDataSetFromMatrix(countData = count_table,
                              colData = sample_table,
                              design = mm)

# Here we get rid of all the genes that are not expressed
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)

# This transforms the counts to have constant variance. This helps ensue that highly or lowly expressed genes will not contribute too much for the PCA or clustering analysis
vsd <- vst(dds, blind = F)

pc_1 <- plot_pca(vsd, group_by = "Sample_name", color_palette = sample_colors)
pc_2 <- plot_pca(vsd, group_by = "group", color_palette = group_colors)
pc_3 <- plot_pca(vsd, group_by = "sex", color_palette = sex_colors)
pc_4 <- plot_pca(vsd, group_by = "mouse", color_palette = mouse_colors)


plot_grid(pc_1, pc_2, pc_3, pc_4,
          labels = c("A", "B", "C", "D"),
          nrow = 2, ncol = 2,
          align = "hv",
          axis = "r",
          label_y = 1,
          label_x = 1)
```


# Differential Expression Analysis {.tabset}


```{r "Make directories", include = F}
# Create the output folders if it doesn't already exist
ifelse(!dir.exists(here(params$output_dir, "DE_files")),
       dir.create(here(params$output_dir, "DE_files"), recursive = TRUE), 
       FALSE)

ifelse(!dir.exists(here(params$output_dir, "images")),
       dir.create(here(params$output_dir, "images")), FALSE)
ifelse(!dir.exists(here(params$output_dir, "images", "heatmaps")),
       dir.create(here(params$output_dir, "images",
                            "heatmaps")), FALSE)
ifelse(!dir.exists(here(params$output_dir, "images", "GSE")),
       dir.create(here(params$output_dir, "images", "GSE")), FALSE)
ifelse(!dir.exists(here(params$output_dir, "GSE_files")),
       dir.create(here(params$output_dir, "GSE_files")), FALSE)
```

```{r "Run differential experession chunks", echo = F}
resultsNames(dds)

dds$sex_group <- paste(dds$sex, dds$group, sep = "_")

sex_group_colors <- c("M_red" = "#EB8B70", "M_green" = "#83CFB0",
                      "F_red" = "#A93A1C", "F_green" = "#2F8060")


# FIX HERE
# Add a sample column and color set option so that the heatmaps can look better
contrasts <- list(
  "red_vs_green" = list(
    "contrast" = c(0, 0, 0, 0, 0, 0, 0, 0, 0.5, 0.5),
    "comparison" = "red:green",
    "sample_column" = "group",
    "color" = group_colors
    ),
  "red_vs_green_male" = list(
    "contrast" = list("sexM.groupred"),
    "comparison" = "M_red:M_green",
    "sample_column" = "sex_group",
    "color" = sex_group_colors
    ),
  "red_green_female" = list(
    "contrast" = list("sexF.groupred"),
    "comparison" = "F_red:F_green",
    "sample_column" = "sex_group",
    "color" = sex_group_colors
    ),
  "red_green_m_vs_f" = list(
    "contrast" = list("sexM.groupred", "sexF.groupred"),
    "comparison" = "red:green",
    "sample_column" = "group",
    "color" = group_colors
    )
  )

DE_chunks <- names(contrasts) %>%
 map(~knit_expand(here("src/scripts/DESeq_template.Rmd"),
                  interaction = TRUE, name = .x))


# DE_chunks <- names(contrasts) %>%
#  map(~knit_expand(here("src/scripts/test.Rmd"),
#                   interaction = TRUE, name = .x))

```

`r knit_child(text = DE_chunks)`

# Gene plots {.tabset}
Gene plots show the normalized expression of key genes across all samples. We can add in any genes that you would like to see in this format.
```{r "Make gene directories", include = F}
ifelse(!dir.exists(here(params$output_dir, "images", "gene_plots")),
       dir.create(here(params$output_dir, "images", "gene_plots")), FALSE)


```

```{r "Run gene experession chunks", echo = F}
sample_column <- "sex_group"
plot_colors <- sex_group_colors

normalized <- FALSE
gene_chunks <- params$genes %>%
  map(~knit_expand(here("src/scripts/gene_plots_template.Rmd")))
```

`r knit_child(text = gene_chunks)`

# Gene plots batch corrected {.tabset}
Gene plots show the normalized and batch corrected expression of key genes across all samples. We can add in any genes that you would like to see in this format.
```{r "Make gene directories normalized", include = F}
ifelse(!dir.exists(here(params$output_dir, "images", "gene_plots_normalized")),
       dir.create(here(params$output_dir, "images", "gene_plots_normalized")),
       FALSE)


```

```{r "Run gene experession chunks normalized", echo = F}
normalized <- TRUE
gene_chunks <- params$genes %>%
  map(~knit_expand(here("src/scripts/gene_plots_template.Rmd")))
```

`r knit_child(text = gene_chunks)`

# Gene plots of the genes with different red and green effects in male and female {.tabset}
These are the genes that came out of the DE for the red vs green in male vs female. These genes should have different red/green effects in the male and female mice


```{r "Run gene plots specific de genes", echo = F}
# Finally check the genes that have unique effects
genes <- read.csv(here("results/R_analysis/DE_files/red_green_m_vs_f.csv"),
                  row.names = 1) %>%
  dplyr::mutate(gene_id = gsub("_ENSMUSG.*", "", rownames(.))) %>%
  dplyr::pull(gene_id)

normalized <- FALSE
gene_chunks <- genes %>%
  map(~knit_expand(here("src/scripts/gene_plots_template.Rmd")))
```

`r knit_child(text = gene_chunks)`

# Session Info

```{r}
sessionInfo()
```

```{r}
ifelse(!dir.exists(here(params$output_dir, "objs")),
       dir.create(here(params$output_dir, "objs")), FALSE)
saveRDS(dds, here(params$output_dir, "objs/combat_obj_obj.rda"))

saveRDS(vsd, here(params$output_dir, "objs/vsd_obj.rda"))


saveRDS(dds, here(params$output_dir, "objs/full_obj.rda"))
```